Learning constraints in tabular data

### 2016-05-13 Paper/ECAI/tabular.tex
chaning minor
Clarifications

\begin{abstract}
%Spreadsheets, comma separated value files and other tabular data representations are in wide use today. However, the correct use and maintenance of functions in and between the tabular data can be error-prone and overwhelming. In this work, we investigate the automatic discovery of constraints (functions and relations) from raw tabular data. Our method takes inspiration form inductive logic programming and constraint satisfiability. We represent common spreadsheet functions as predicates with arguments that must satisfy number of constraints over the arguments. This allows generic constraint satisfaction techniques to be used to find all such satisfying predicates/functions within the rows, columns and submatrices of the tables. We show the effectiveness and accuracy of this approach on a number of spreadsheets from varying sources.
Spreadsheets, CSV (comma separated value) files and other tabular data representations are in wide use today. However, modeling, maintaining and discovering formulas in tabular data and between spreadsheets can be time consuming and error-prone. In this work, we investigate the automatic discovery of constraints (functions and relations) from raw tabular data. We see multiple promising applications for this technique, e.g. in rediscovering constraints, auto-completion and error checking. Our method takes inspiration from inductive logic programming, constraint learning and constraint satisfaction. Common spreadsheet functions are represented as predicates whose arguments must satisfy a number of constraints. Constraint learning techniques are used to identify predicates (constraints) that hold between the rows, columns and submatrices of the tables. New types of constraints can be easily added to the system by specifying them in a declarative language. We show that our approach is able to accurately discover constraints in a number of spreadsheets from varying sources.
\end{abstract}

@@

In this paper,  we investigate whether learning techniques can be used to infer constraints (formula's and other relations) from raw spreadsheet data.
This is a new and unconventional machine learning problem.
Consider the example in Figure~\ref{fig:main_example}, where the header's names already suggest the usage of spreadsheet operations such as \textit{average} or \textit{sum}.
Looking at the first row in Table~2 and the data in Table~1, it is clear that the values in this row were computed by summing the values in the corresponding column of Table~1.
Examining Table~1 also shows that computations are not only performed column-wise but row-wise as well: the cells in the column \textbf{Total} are obtained by taking the row-wise sum of the previous four columns.
This already provides a flavor of how this problem is different from standard data mining settings, where the data is just in rows and variables are in columns. Here everything is mixed. The data is relational on the one hand, since we have multiple table with relationships between them. And on the other hand, the data is mixed textual and numeric, since people tend to use spreadsheets for financial and accounting computations. One can see that understanding and learning constraints in tabular data is hence a new and challenging problem for machine learning. Furthermore, it is not only challenging for a machine but for humans as well, e.g. the influential Reinhart-Rogoff economical paper ``Growth in a Time of Debt'' turned to be bogus due an excel formula mistake \cite{flaw_excel}.

@@

The question that we answer in this paper is: is it possible to discover or reconstruct structural constraints (relations, functions) in flat tabular spreadsheet data?
To answer this question we contribute a general-purpose algorithm and the system, named \sname (pronounced as the word ``tackle''), for discovering such constraints. Furthermore, the algorithm accepts declarative specifications of the constraints and predicates of interest (the "bias") and covers both textual and numeric fields. It works in an unsupervised setting, as only the tabular data is provided and no examples need to be specified, and discovers column- and row-wise constraints.
We demonstrate the utility of our approach in an experimental evaluation.
Moreover, we sketch several application scenario's such as autocompletion and error detection.
\\\\
This paper is organized as follows.
Section \ref{sec:formalization} introduces concepts relevant to our approach. Section \ref{sec:problem_statement} presents the problem statement. Section \ref{sec:approach} describes the approach. Section \ref{sec:evaluation} presents the evaluation of the approach. Section \ref{sec:applications} shows how our system can be used for applications. Section \ref{sec:related_work} presents an overview of the related work. Section \ref{sec:conclusions} provides conclusions.

@@

\subsection{Cells, vectors and groups}
A spreadsheet may conceptually consist of multiple tables, such as in Figure~\ref{fig:main_example}. We will describe a method of extracting tables from a big spreadsheet in Section~\ref{sec:approach}, however this is not the core of our learning method. Our formalization reasons over tables directly.

Formally, a table is an $n \times m$ matrix. Each entry is called a \textit{cell}.
A cell has a {\em type}, which can be numeric or textual. We further distinguish numeric types in subtypes: integer and float. We also consider \textit{None} as a special type when a cell is empty; \textit{None} is a subtype of all other types.
A set of cells is called \textit{type-consistent} iff all cells in the set are of the same type (\textit{None} is consistent with everything).
%Certain constraints, such as \textit{rank} or \textit{series}, make use of the numeric subtype by requiring its arguments to be integers.

@@

  \item
  A \textbf{subgroup} $G'$ of a group $G$ is simply a projection of $G$ onto a consecutive subset its vectors. % is a subrange of vectors belonging to a group $G$. If the subgroup $g$ must contain only one vector, we write $g \in G$. %It defines what it means for a constraint to be well-formed
%We refer to a subgroup of a group $G$ in \groups as $g$. A \textit{subgroup} $g$ is a subrange of vectors in a group $G$. If a subgroup $g$ must contain only one vector, we write $g \in G$.
% A \textit{subgroup} $G'$ is a subrange of vectors belonging to a group $G$. If the subgroup $\{v\}$ must contain only one vector, we write $v \in G$.
\end{itemize}
%For notational convenience, we will refer to vectors in {\rm roman} and to groups in {\bf bold}. {\bf CHECK}

@@

\item the \textit{\CSignature} defines the properties that the arguments of the predicate must satisfy. That is, the required types of the arguments, together
with possible further restrictions, e.g., on their length. For example, the signature can restrict the types of the arguments to be groups of integers and require that the length of all arguments be equal. In terms of logical and relational learning \cite{luc_book}, such a \CSignature is known as the {\em bias} of the learner, it specifies when the constraint is well-formed.

@@

\paragraph{Example}
Several example constraints are illustrated in Table 1.  For example, the constraint template \textit{rank} has
\begin{itemize}
  \item \CName: \ecrank{G_r}{G_x}, where $G_r$ and $G_x$ are the arguments;
\item \CSignature: the two arguments $G_r$ and $G_x$ must be numeric groups and of size one; furthermore, the group $G_r$ must be integer (the rank) and the length of groups $G_x$ and $G_r$ must be identical, i.e., $length(G_r) = length(G_x)$;
\item \CFunction: \ecrank{G_r}{G_x}, whenever each value of $G_r$ is the rank of the corresponding value in $G_x$.
% is satisfied in the mapping $\{ v_x \mapsto g_x, v_y \mapsto g_y \}$ satisfies the constraint iff $g_x \in G_x, g_y \in G_y$ ($G_x \in \groups, G_y \in \groups$) and each value in $g_y$ is the rank of the corresponding value in $g_x$ (possibly with ties).
\end{itemize}

When looking into constraints like \ecrank{G_r}{G_x}, it is helpful to see the analogy with both first order logic (FOL) and with constraint satisfaction problems.
From a FOL perspective, the name of the constraint (\textit{RANK}) is just the predicate, and the arguments $G_r$ and $G_x$ are the terms, which can be seen as either uninstantiated variables or as values (concrete groups with values).
This also holds in our setting: when we write
      $\ecrank{\range{T}{\rangeall}{8}}{\range{T}{\rangeall}{7}}$,
we can interpret the argument ${\range{T}{\rangeall}{8}}$ as a group variable that would apply to any table $T$ of the right dimensions.
So, $T$ is viewed as table variable here, with the effect that  ${\range{T}{\rangeall}{8}}$ is a group variable.

\tias{Of the right (minimum) dimension?? Note that this is also sloppy because the ':8' is not considered part of the variable...},
Alternatively, and perhaps more naturally, we can interpret the argument as the value it takes in a particular table $T$ such as $T_1$ in Figure 1, that is,
${\range{T_1}{\rangeall}{8}} =  [1,2,3,4]$.
Then we can also talk about constraints being satisfied. Basically, a constraint $c(G_1, ..., G_n)$ is {\em satisfied}  in a set of tables ${\cal T}$ if and only
if the constraint holds on the values that the $A_i$ take in the set of tables ${\cal T}$. \tias{I find this confusing: a group as we defined it is defined on one table, not on multiple tables. we should separate the table variable from the 'range' variables then...}
%Figure 3 provides a list of constraints that are satisfied by the tables in Figure 1.

@@

Two constraints are \textit{equivalent} if they imply one another. %More formally, $\forall G'_1, ..., G'_n$ s.t. $\{G''_1, ..., G''_n\} = \{G'_1, ..., G'_n\},\, c_1(G'_1, ... G'_n) \leftrightarrow c_2(G''_1, ..., G''_n)$.
An example within the same \template is for the sum over vectors constraint: $G_1 = +(G_2, G_3) \leftrightarrow G_1 = +(G_3, G_2)$. A related example over multiple constraints is $G_1 = +(G_2, G_3) \leftrightarrow G_1 = \textit{SUM}(H)$ with $\textit{size}(H) = 2$ and $G_2 \subset H, G_3 \subset H$.
\luc{Do we really need this, or can we write it in a real equivalence --- conjunctive formula?}

@@

\tias{I am not defining a new problem for 'most specific canonical tabular constraint learning' because we loosely showed how implications and equivalences would be expressed, but not how to express canonicity and we did not claim that most specific is always best.}
\tias{Note also that our method finds ALL constraints; it uses specificity to faster traverse the search space, but the resulting set of constraints found would be the same, whether or not using the specificity (see 4.2.1).}

@@

From each table, row-oriented and column-oriented groups are then extracted by partitioning the rows (columns) into type-consistent parts. This step can also be verified by a user in the visual tool.

@@

\paragraph{Application specific system extension}
\sergey{under question now, might worth deleting} \\
\sergey{Explain here, how the system can be extended with a new constraints like vr = floor(v), float(v), int(vr), len(v)=len(vr), v,vr are single vectors}

>>>>

\begin{abstract}
%Spreadsheets, comma separated value files and other tabular data representations are in wide use today. However, the correct use and maintenance of functions in and between the tabular data can be error-prone and overwhelming. In this work, we investigate the automatic discovery of constraints (functions and relations) from raw tabular data. Our method takes inspiration form inductive logic programming and constraint satisfiability. We represent common spreadsheet functions as predicates with arguments that must satisfy number of constraints over the arguments. This allows generic constraint satisfaction techniques to be used to find all such satisfying predicates/functions within the rows, columns and submatrices of the tables. We show the effectiveness and accuracy of this approach on a number of spreadsheets from varying sources.
Spreadsheets, CSV (comma separated value) files and other tabular data representations are in wide use today. However, modeling, maintaining and discovering formulas in tabular data and between spreadsheets can be time consuming and error-prone. In this work, we investigate the automatic discovery of constraints (functions and relations) from raw tabular data. We see multiple promising applications for this technique, e.g. in rediscovering constraints, auto-completion and error checking. Our method takes inspiration from inductive logic programming, constraint learning and constraint satisfaction. Common spreadsheet functions are represented as predicates whose arguments must satisfy a number of constraints. Constraint learning techniques are used to identify predicates (constraints) that hold between the rows, columns and submatrices of the tables. New types of constraints can be easily added to the system by specifying them using in a declarative language. We show that our approach is able to accurately discover constraints in a number of spreadsheets from varying sources.
\end{abstract}

@@

In this paper,  we investigate whether learning techniques can be used to infer constraints (formula's and other relations) from raw spreadsheet data.
This is a new and unconventional machine learning problem.
Consider the example in Figure~\ref{fig:main_example}, where the header's names already suggest the usage of spreadsheet operations such as \textit{average} or \textit{sum}.
Looking at the first row in Table~2 and the data in Table~1, it is clear that the values in this row were computed by summing the values in the corresponding column of Table~1.
Examining Table~1 also shows that computations are not only performed column-wise but row-wise as well: the cells in the column \textbf{Total} are obtained by taking the row-wise sum of the previous four columns.
This already provides a flavor of how this problem is different from standard data mining settings, where the data is just in rows and variables are in columns. Here everything is mixed. The data is relational on the one hand, since we have multiple table with relationships between them. And on the other hand, the data is mixed textual and numeric, since people tend to use spreadsheets for financial and accounting computations. One can see that understanding and learning constraints in tabular data is hence a new and challenging problem for machine learning. Furthermore, it is not only challenging for a machine but for humans as well.

@@

The question that we answer in this paper is: whether it is possible to discover or reconstruct structural constraints (relations, functions) in flat tabular spreadsheet data?
To answer this question we contribute a general-purpose algorithm and system for discovering such constraints. Furthermore, the algorithm accepts declarative specifications of the constraints and predicates of interest (the "bias") and covers both textual and numeric fields. It works in an unsupervised setting, as only the tabular data is provided and no examples need to be specified, and discovers column- and row-wise constraints.
We demonstrate the utility of our approach in an experimental evaluation.
Moreover, we sketch several application scenario's, such as autocompletion and error detection.
\\\\
This paper is organized as follows.
Section \ref{sec:formalization} introduces concepts relevant to our approach. Section \ref{sec:problem_statement} presents the problem statement. Section \ref{sec:approach} describes the approach. Section \ref{sec:evaluation} presents the evaluation of the approach. Section \ref{sec:applications} shows how our system can be used for applications such as autocompletion and error detection. Section \ref{sec:related_work} present an overview of the related work. And Section \ref{sec:conclusions} provides conclusions.

@@

\subsection{Cells, vectors and groups}
A spreadsheet may conceptually consist of multiple tables, such as in Figure~\ref{fig:main_example}. We will describe a method of extracting tables from a big spreadsheet in Section~\ref{todo}, however this is not the core of our learning method. Our formalization reasons over tables directly.

Formally, a table is an $n \times m$ matrix. Each entry is called a \textit{cell}.
A cell has a {\em type}, which can be numeric or textual. We further distinguish numeric types in subtypes integer and float. We also consider \textit{None} as a special type when a cell is empty; \textit{None} is a subtype of all other types.
A set of cells is called \textit{type-consistent} iff all cells in the set are of the same type.
%Certain constraints, such as \textit{rank} or \textit{series}, make use of the numeric subtype by requiring its arguments to be integers.

@@

  \item
  A \textbf{subgroup} $G'$ of a group $G$ is simply a projection of $G$ onto a consecutive subset of its vectors. % is a subrange of vectors belonging to a group $G$. If the subgroup $g$ must contain only one vector, we write $g \in G$. %It defines what it means for a constraint to be well-formed
%We refer to a subgroup of a group $G$ in \groups as $g$. A \textit{subgroup} $g$ is a subrange of vectors in a group $G$. If a subgroup $g$ must contain only one vector, we write $g \in G$.
% A \textit{subgroup} $G'$ is a subrange of vectors belonging to a group $G$. If the subgroup $\{v\}$ must contain only one vector, we write $v \in G$.
\end{itemize}
%For notational convenience, we will refer to vectors in {\rm roman} and to groups in {\bf bold}. {\bf CHECK}

@@

\item the \textit{\CSignature} defines the properties that the arguments of the predicate must satisfy. That is, the required types of the arguments, together
with possible further restrictions, e.g., on their length. For example, the signature can restrict the types of the arguments to be groups of integers and require that the length of all arguments be equal. \\
In terms of logical and relational learning \cite{luc_book}, such a \CSignature is known as the {\em bias} of the learner, it specifies when the constraint is well-formed.

@@

\paragraph{Example}
Several example constraints are illustrated in Table 1.  For example, the constraint template \textit{rank} has
\begin{itemize}
\item \CName: \textit{$B$ = RANK($A$)}, where $B$ and $A$ are the arguments;
\item \CSignature: the two arguments $B$ and $A$ must be numeric groups and of size 1; furthermore, the group $B$ must be integer (the rank) and the length of groups $A$ and $B$ must be identical, i.e., $length(B) = length(A)$;
\item \CFunction: \textit{$B$ = RANK($A$)} whenever each value of $B$ is the rank of the corresponding value in $A$.
% is satisfied in the mapping $\{ v_x \mapsto g_x, v_y \mapsto g_y \}$ satisfies the constraint iff $g_x \in G_x, g_y \in G_y$ ($G_x \in \groups, G_y \in \groups$) and each value in $g_y$ is the rank of the corresponding value in $g_x$ (possibly with ties).
\end{itemize}

When looking into constraints like \textit{$B$ = RANK($A$)}, it is helpful to see the analogy with both first order logic (FOL) and with constraint satisfaction problems.
From a FOL perspective, the name of the constraint ($\mathit{RANK}$) is just the predicate, and the arguments $B$ and $A$ are the terms, which can be seen as either uninstantiated variables or as values (concrete groups with values).
This also holds in our setting: when we write
      $\ecrank{\range{T}{\rangeall}{8}}{\range{T}{\rangeall}{7}}$,
we can interpret the argument ${\range{T}{\rangeall}{8}}$ as a group variable that would apply to any table $T$ (provided~$T$ has an 8th column).
So, $T$ is viewed as table variable here, with the effect that  ${\range{T}{\rangeall}{8}}$ is a group variable.

\samuel{\tias{Of the right (minimum) dimension?? Note that this is also sloppy because the ':8' is not considered part of the variable...} True, I changed the wording a bit but personally I would leave it at this now, you can always say you look at $T[:,i:j]$ iff $cols(T) >= max(i,j)$}.
Alternatively, and perhaps more naturally, we can interpret the argument as the value it takes in a particular table $T$ such as $T_1$ in Figure 1, that is,
${\range{T_1}{\rangeall}{8}} =  [2,1,3,4]$.
Then we can also talk about constraints being satisfied. Basically, a constraint $c(A_1, ..., A_n)$ is {\em satisfied}  in a set of tables ${\cal T}$ if and only if the constraint holds on the values that the $A_i$ take in the set of tables ${\cal T}$. \samuel{\tias{I find this confusing: a group as we defined it is defined on one table, not on multiple tables. we should separate the table variable from the 'range' variables then...} waiting for Luc's opinion here}
%Figure 3 provides a list of constraints that are satisfied by the tables in Figure 1.

@@

Two constraints are \textit{equivalent} if they imply one another. %More formally, $\forall G'_1, ..., G'_n$ s.t. $\{G''_1, ..., G''_n\} = \{G'_1, ..., G'_n\},\, c_1(G'_1, ... G'_n) \leftrightarrow c_2(G''_1, ..., G''_n)$.
An example within the same \template is for the sum over vectors constraint: $G_1 = +(G_2, G_3) \leftrightarrow G_1 = +(G_3, G_2)$. A related example over multiple constraints is $G_1 = +(G_2, G_3) \leftrightarrow G_1 = \textit{SUM}(H)$ with $\textit{size}(H) = 2$ and $G_2 \subset H, G_3 \subset H$.
\luc{Do we really need this, or can we write it in a real equivalence --- conjunctive formula?}
\samuel{Tias, Sergey what do you think? Also I find this notation confusing: +(G2, G3), why not just use G1 = G2 + G3?}

@@

%\tias{I am not defining a new problem for 'most specific canonical tabular constraint learning' because we loosely showed how implications and equivalences would be expressed, but not how to express canonicity and we did not claim that most specific is always best.}
%\tias{Note also that our method finds ALL constraints; it uses specificity to faster traverse the search space, but the resulting set of constraints found would be the same, whether or not using the specificity (see 4.2.1).}

@@

Row-oriented and column-oriented groups are then extracted from each table by partitioning the rows (columns) into type-consistent parts.

@@

\paragraph{Application specific system extension}
\sergey{Explain here, how the system can be extended with a new constraints like vr = floor(v), float(v), int(vr), len(v)=len(vr), v,vr are single vectors}
